Exploring the Role of Feedback in Machine Learning Jupyter Notebooks

### 2024-07-02 report/emse24.tex
chore(report): emse24 migrate feedback from Luis

In software engineering, tests are employed to document the expected behaviour of code. Failing tests indicate when these expectations are no longer met. Similarly, assertions in Jupyter notebooks can serve as documentation for the decisions made during model development. Embedding assertions in the notebook can ensure that any deviations from expected data conditions or model performance are promptly detected, thus maintaining the integrity of the ML pipeline.

@@

% NOTE pimentel2019large-scale:
% NOTE kery2018story
% NOTE rule2018exploration

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{nb.pdf}
  \label{fig:nb}
  \caption{Example Jupyter notebook with code and markdown cells, adapted from~\cite[Figure~1]{pimentel2019large-scale}.}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[width=0.75\linewidth]{data-collection.pdf}
  \label{fig:data-collection}
  \caption{Overview of data collection methodology used in this study}
\end{figure}

@@

We check the validity of the underlying JSON structure in each notebook using the nbformat tool\footnote{https://nbformat.readthedocs.io/en/latest/}. Since the focus of this study is on the analysis of Python code, we only include notebooks that contain at least one code cell. Finally, to focus on Python code written specifically for ML projects, we only include notebooks that import popular ML libraries~\footnote{These libraries are derived from white literature---https://www.coursera.org/articles/python-machine-learning-library, https://www.geeksforgeeks.org/best-python-libraries-for-machine-learning, https://www.datacamp.com/blog/top-python-libraries-for-data-science, https://www.kdnuggets.com/2020/11/top-python-libraries-data-science-data-visualization-machine-learning.html, https://lp.jetbrains.com/python-developers-survey-2022}.

@@

Similarly, for implicit feedback, we first isolate code cells that produce an output. The output of a cell may either be from a ``print'' statement or the last statement of the cell. We collect both of them using the AST of the cell.

@@

% TODO: might need a visual aid here

We allocated a fixed time resource of 100 hours to conduct the case study analysis of all candidates (assertions and cell outputs). To surface interesting candidates for the case studies, we apply NLP techniques as described below.

@@

% TODO: feedback from Diomidis
% - Sections 4.1 and 4.2 are very interesting but overly detailed. I suggest to move most of the content to an appendix or an online appendix, and only present the uses in a table or a taxonomy tree.  In addition, consider moving the more general observations into a discussion section.
% - Consider starting with RQ2 (making it RQ1), because assertions are a concrete feature of Python code.  Then move to RQ1, which deals with implicit assertions.
% - I am afraid I cannot understand RQ3; consider revising it.  I would think that visualizations are verified visually, rather than through assertions.  You may have encountered some parts, but I would assume that the assertions mostly deal with things that can be more accurately and effectively be expressed as a predicate rather than viewed in a chart.
% TODO: some sort of introduction to the following sub-sections here; we need to explain that for code cell outputs we try to show the python statement that produces the output along with the actual output. Somethimes actual output is not possible (eg. P68) because the statement may be inside an if statement.

\subsection{(RQ1) What are the characteristics of feedback mechanisms in ML Jupyter notebooks?}

@@

\subsection{(RQ2) How is feedback from the output of code cells used when writing ML code in Jupyter notebooks?}~\label{sec:result-rq1}

@@

Table~\ref{tab:distribution-check} shows examples of outputs used to check the distribution of the data. In output \emph{O2} the distribution of feature \texttt{Age} is checked against the two labels (\texttt{Survived} and \texttt{Died}) of the target feature. Output \emph{O9} presents another example where the distribution of a continuous feature \texttt{likes} is checked with-respect-to all the values of another categorical feature \texttt{category\_id}. Output \emph{O14} produces a pandas dataframe which is used to check the mean value of the features \texttt{Age}, \texttt{SibSp}, \texttt{Parch} and \texttt{Fare} against each label in the target feature \texttt{Survived}. Output \emph{O48} produces a pandas dataframe with the descriptive statistics of all features in the training data. The statement is written immediately after normalising the features in the training data. Hence, the user manually verifies if the normalisation was successful by looking at the mean and standard deviation of the features in the output. Similarly, \emph{O25} is written after discretizing a continuous feature \texttt{OverallQual} using binning.

@@

\highlight{\textbf{Finding 1.1} The study identifies seven outputs used to check data distribution, including visualisations like categorical plots, KDE plots, and count plots (O2, O9 and O25), and statistical summaries from pandas dataframes (O14 and O48). These checks are used for making informed decisions about data transformations such as scaling, normalizing and handling outliers, ensuring data integrity and improving model training reliability during the EDA phase.}

@@

Linear machine learning models achieve optimal performance when the target variable can be expressed as a linear combination of the input features. However, features within a dataset that exhibit a linear relationship are considered redundant as they convey the same information to the model during training. Consequently, the feature engineering stage often involves removing such features to create a more efficient training dataset~\cite{shome2022data}.

\highlight{\textbf{Finding 1.2} The study identifies two outputs used to verify linear relationships between features in the data. visualisations such as a scatter plot with a regression line (O10) and a line plot (O6) are employed to assess the linearity between features. These checks are used for identifying and potentially removing redundant features or performing feature selection to ensure that the training dataset is optimized for linear ML models.}

@@

% NOTE: O107 is a print statement which is execute even if the prior statement that loads the model fails...
Table~\ref{tab:resource-check} presents outputs used to verify availability of resources on the system in which the notebook is being executed. Outputs \emph{P68}, \emph{P82} and \emph{P86} check for the availability of compute resources, such as a GPU or a TPU. This study also finds outputs used to verify that a dataset or a pre-trained model has been successfully loaded into memory. For example, Output \emph{P107} is written after loading a pre-trained model is loaded into memory. Output \emph{O64} generates a pandas dataframe with the last five rows of the data while \emph{O66} shows the shape of the data. These statements are written directly after loading the data from the file system to manually verify that the data was loaded without errors.

Additionally, print statement \emph{P71} is used to ensure that a certain version of an external library is present on the system. This is interesting since external dependencies in Python projects are typically managed using dependency management solutions such as a \emph{requirements.txt}\footnote{https://pip.pypa.io/en/stable/reference/requirements-file-format/} file or external tools such as \emph{pipenv}\footnote{https://pipenv.pypa.io/en/latest/} and \emph{poetry}~\footnote{https://python-poetry.org/}.

@@

\subsubsection{Model performance check ($N = 33$)}
% (O3, O50, O52, O57, O74, P3, P6, P8--12, P16--19, P23, P24, P28, P30, P34, P42, P47, P48, P50, P51, P54, P55, P57, P58, P61, P78, P93)

\begin{table}
  \centering
  \caption{Various cell outputs used to check the performance of an ML model after training.}
  \begin{tabular}{@{}m{0.05\textwidth} m{0.4\textwidth} m{0.4\textwidth}@{}}
    \toprule
    \emph{\textbf{Key}}&
    \emph{\textbf{Code}}&
    \emph{\textbf{Visual Output}}\\
    \midrule

@@

\begin{table}
  \centering
  \caption{Example of outputs used to monitor model training.}
  \begin{tabular}{@{}m{0.05\textwidth} m{0.8\textwidth}@{}}
    \toprule
    \emph{\textbf{Key}}&
    \emph{\textbf{Code}}\\
    \midrule

@@

Studies with human subjects have been conducted to gain a deeper understanding of the challenges faced by ML practitioners when developing ML models inside notebooks. The results indicate that ML practitioners work in a highly iterative fashion, often experimenting with multiple strategies to analyse the data or produce meaningful visualisations~\cite{kandel2012enterprise, kery2018story, liu2019understanding, chattopadhyay2020whats}. Studies have also been conducted to understand how practitioners generate, evaluate and manage alternative hypothesis, visual designs, methods, tools, algorithms and data sources to arrive at the final implementation~\cite{liu2019understanding,kandel2012enterprise}. The findings from these studies can be used as guidelines for improving existing notebook technologies or designing new ones.

To manage and prune multiple versions of code that accumulate over time when developing in notebooks, \emph{Head et al} propose a code gathering tool that allows practitioners to review and only keep the relevant version of code. Other tools such as \textit{WrangleDoc} and \textit{VizSmith} have also been proposed to aid ML practitioners when working within computational notebooks~\cite{yang2021subtle, bavishi2021vizsmith}.

@@

To test the correctness of ML models, several improvements over existing test adequacy metrics have been proposed. Tools such as \textit{DeepXplore} and \textit{DeepGauge} propose new test adequacy metrics such as neuron coverage adapted for ML enabled software systems~\cite{pei2017deepxplore, ma2018deepgauge, gerasimou2020importance-driven}. Formal verification methods have also been proposed that try to provide formal guarantee of robustness against adversarial examples. Such methods however are only feasible for statistical ML models and become computationally expensive for more complex models such as deep neural networks~\cite{zhu2021deepmemory, baluta2021scalable}. Several works have been conducted on generating and detecting adversarial inputs for ML models. Data augmentation techniques based on fuzzing, search based software testing and mutation testing have been proposed to generate adversarial examples that can be used during model training to improve its robustness~\cite{braiek2019deepevolution, gao2020fuzz, wang2021robot, zhang2020white-box}. It is however not possible to include all variations of adversarial examples into the training data. Thus, methods have been proposed to detect adversarial inputs during runtime~\cite{xiao2021self-checking, wang2020dissector, wang2019adversarial, berend2020cats}.

@@

\textit{ESCAPE}, \textit{GAM Coach}, \textit{Angler} and \textit{Drava} are a few other tools that have been proposed to handle specialised use-cases such as identifying systematic errors in ML models, generating counterfactual explanations for Generalised Additive Models, prioritising machine translation model improvements and relating human concepts with semantic dimensions extracted by ML models during disentangled representation learning\cite{ahn2023escape, wang2023gam, robertson2023angler, wang2023drava}.

>>>>

% TODO: feedback from Luis
% - not flowing with previous paragraph
% - The previous paragraph could mention something like ad-hoc manual tests. And then in this paragraph we could say that software engineering proposes practices on automated testing that can be useful in this context
% - in theory, the main goal of tests is to verify the requirements of the software. Documentation is a by product. (I like the idea in this paragraph but we dont want to touch a raw nerve from reviewers)
In software engineering, tests are employed to document the expected behaviour of code. Failing tests indicate when these expectations are no longer met. Similarly, assertions in Jupyter notebooks can serve as documentation for the decisions made during model development. Embedding assertions in the notebook can ensure that any deviations from expected data conditions or model performance are promptly detected, thus maintaining the integrity of the ML pipeline.

@@

% NOTE pimentel2019large-scale
% NOTE kery2018story
% NOTE rule2018exploration

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{nb.pdf}
  \caption{Example Jupyter notebook with code and markdown cells, adapted ~\cite[Figure~1]{pimentel2019large-scale}.}
  \label{fig:nb}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[width=0.75\linewidth]{data-collection.pdf}
  \caption{Overview of data collection methodology used in this study}
  \label{fig:data-collection}
\end{figure}

@@

% TODO: feedback from Luis
% - explain why we check the validity
% - we cannot guarantee that the white literature will be online in the next months/years and make it difficult to collect an important detail for reproducibility. I would add to the text at least how many libraries we added, give a few examples and point to our replication package,
We check the validity of the underlying JSON structure in each notebook using the nbformat tool~\footnote{https://nbformat.readthedocs.io/en/latest/}. Since the focus of this study is on the analysis of Python code, we only include notebooks that contain at least one code cell. Finally, to focus on Python code written specifically for ML projects, we only include notebooks that import popular ML libraries~\footnote{These libraries are derived from grey literature---https://www.coursera.org/articles/python-machine-learning-library, https://www.geeksforgeeks.org/best-python-libraries-for-machine-learning, https://www.datacamp.com/blog/top-python-libraries-for-data-science, https://www.kdnuggets.com/2020/11/top-python-libraries-data-science-data-visualization-machine-learning.html, https://lp.jetbrains.com/python-developers-survey-2022}.

@@

% TODO: feedback from Luis
% - only here, a reader start understanding that we want to collect assert, prints, and lasts. Perhaps we want to make that clear in the beginning of this subsection or in the one above it
Similarly, for implicit feedback, we first isolate code cells that produce an output. The output of a cell may either be from a ``print'' statement or the last statement of the cell. We collect both of them using the AST of the cell.

@@

% TODO: might need a visual aid here
% TODO: feedback from Luis
% - motivate this approach
% - not sure we want to say NLP here. I would call it text processing. NLP kinda implies that we are automatically interpreting the text
We allocated a fixed time resource of 100 hours to conduct the case study analysis of all candidates (assertions and cell outputs). To surface interesting candidates for the case studies, we apply NLP techniques as described below.

@@

% TODO: feedback from Diomidis
% - Sections 4.1 and 4.2 are very interesting but overly detailed. I suggest to move most of the content to an appendix or an online appendix, and only present the uses in a table or a taxonomy tree.  In addition, consider moving the more general observations into a discussion section.
% - Consider starting with RQ2 (making it RQ1), because assertions are a concrete feature of Python code.  Then move to RQ1, which deals with implicit assertions.
% - I am afraid I cannot understand RQ3; consider revising it.  I would think that visualizations are verified visually, rather than through assertions.  You may have encountered some parts, but I would assume that the assertions mostly deal with things that can be more accurately and effectively be expressed as a predicate rather than viewed in a chart.
% TODO: some sort of introduction to the following sub-sections here; we need to explain that for code cell outputs we try to show the python statement that produces the output along with the actual output. Somethimes actual output is not possible (eg. P68) because the statement may be inside an if statement.
\subsection{(RQ1) What are the characteristics of feedback mechanisms in ML Jupyter notebooks?}

@@

\subsection{(RQ2) How is feedback from the output of code cells used when writing ML code in Jupyter notebooks?}~\label{sec:result-rq1}
% TODO: feedback from Luis
% -  up to this point I miss how exactly we plan to answer this question. Reading the results I get that we are answering by clustering outputs together based on their respective ML/DS task and then by analysing how that task is accomplished. We need to explain this either in the beginning of section 4.1 or in section 3

@@

% TODO: feedback from Luis
% - I miss a bit the link between the RQ and the different subsections 4.1.1
% -  I also miss an explanation of the table. what is O2? We should have a sentence explaining that each of the examples in the table are identified by a key,  the respective code and visual output, if any.
% - Explain also how the figure is extracted. This is important because it is the first one. For the next ones we can just say that we follow the same structure.
% - Another alternative would be to explain this generally in the intro of section 4.1 and then follow it consistently.
% - I also miss how the examples are extracted or what they mean here.
% - I mean, are these all the examples we found for distribution check? If so, make it clear.
% - add: "with key" to be closer to the table structure
Table~\ref{tab:distribution-check} shows examples of outputs used to check the distribution of the data. In output \emph{O2} the distribution of feature \texttt{Age} is checked against the two labels (\texttt{Survived} and \texttt{Died}) of the target feature. Output \emph{O9} presents another example where the distribution of a continuous feature \texttt{likes} is checked with-respect-to all the values of another categorical feature \texttt{category\_id}. Output \emph{O14} produces a pandas dataframe which is used to check the mean value of the features \texttt{Age}, \texttt{SibSp}, \texttt{Parch} and \texttt{Fare} against each label in the target feature \texttt{Survived}. Output \emph{O48} produces a pandas dataframe with the descriptive statistics of all features in the training data. The statement is written immediately after normalising the features in the training data. Hence, the user manually verifies if the normalisation was successful by looking at the mean and standard deviation of the features in the output. Similarly, \emph{O25} is written after discretizing a continuous feature \texttt{OverallQual} using binning.

@@

% TODO: Feedback from Luis
% - I think you should mention the fact that there are both prints and visualisations for this task, the fact that the approach can be different not only depending on the type of data,  but also depending on the preferences of the developers (maybe we cannot really claim this last part but at least mention that there are multiple visualisations for the same scenario and developers simply need to opt for one.
\highlight{\textbf{Finding 1.1} The study identifies seven outputs used to check data distribution, including visualisations like categorical plots, kernel density estimate plots and count plots (\emph{O2}, \emph{O9} and \emph{O25}), and statistical summaries from pandas dataframes (\emph{O14} and \emph{O48}). These checks are used for making informed decisions about data transformations such as scaling, normalising and handling outliers, ensuring data integrity and improving model training reliability during the exploratory data analysis phase.}

@@

% TODO: feedback from Luis
% - this explains what the code does but it does not pinpoint observations on how the output of code cells are used. E.g., here it is mostly visualisations - that's an observation.  They are both scatterplots, one had to change the height of the plot (not sure that's relevant. We do not observe prints or asserts that help with this task.
Linear machine learning models achieve optimal performance when the target variable can be expressed as a linear combination of the input features. However, features within a dataset that exhibit a linear relationship are considered redundant as they convey the same information to the model during training. Consequently, the feature engineering stage often involves removing such features to create a more efficient training dataset~\cite{shome2022data}.

\highlight{\textbf{Finding 1.2} The study identifies two outputs used to verify linear relationships between features in the data. visualisations such as a scatter plot with a regression line (\emph{O10}) and a line plot (\emph{O6}) are employed to assess the linearity between features. These checks are used for identifying and potentially removing redundant features or performing feature selection to ensure that the training dataset is optimized for linear ML models.}

@@

% NOTE: O107 is a print statement which is execute even if the prior statement that loads the model fails...
% TODO: feedback from luis
% - I probably missed it but it should be clear why we have P and O in the key, given that they are all outputs.
Table~\ref{tab:resource-check} presents outputs used to verify availability of resources on the system in which the notebook is being executed. Outputs \emph{P68}, \emph{P82} and \emph{P86} check for the availability of compute resources, such as a GPU or a TPU. This study also finds outputs used to verify that a dataset or a pre-trained model has been successfully loaded into memory. For example, Output \emph{P107} is written after loading a pre-trained model is loaded into memory. Output \emph{O64} generates a pandas dataframe with the last five rows of the data while \emph{O66} shows the shape of the data. These statements are written directly after loading the data from the file system to manually verify that the data was loaded without errors.

Additionally, print statement \emph{P71} is used to ensure that a certain version of an external library is present on the system. This is interesting since external dependencies in Python projects are typically managed using dependency management solutions such as a \emph{requirements.txt}~\footnote{https://pip.pypa.io/en/stable/reference/requirements-file-format/} file or external tools such as \emph{pipenv}~\footnote{https://pipenv.pypa.io/en/latest/} and \emph{poetry}~\footnote{https://python-poetry.org/}.

@@

\subsubsection{Model performance check ($N = 33$)}
% (O3, O50, O52, O57, O74, P3, P6, P8--12, P16--19, P23, P24, P28, P30, P34, P42, P47, P48, P50, P51, P54, P55, P57, P58, P61, P78, P93)
% TODO: feedback from Luis
% - if needed, we can merge Excution time check here. (both execution time and model correctness can be presented as a KPI)
\begin{table}
  \centering
  \caption{Various cell outputs used to check the performance of an ML model after training.}
  \begin{tabular}{@{}m{0.05\textwidth} m{0.4\textwidth} m{0.4\textwidth}@{}}
    \toprule
    \emph{\textbf{Key}}&
    \emph{\textbf{Code}}&
    \emph{\textbf{Visual Output}}\\
    \midrule

@@

% TODO: feedback from luis
% - this cluster makes me wonder: should we make a point that some of these outputs may happen by chance and were not explicitly defined by users/developers?
\begin{table}
  \centering
  \caption{Example of outputs used to monitor model training.}
  \begin{tabular}{@{}m{0.05\textwidth} m{0.8\textwidth}@{}}
    \toprule
    \emph{\textbf{Key}}&
    \emph{\textbf{Code}}\\
    \midrule

@@

Studies with human subjects have been conducted to gain a deeper understanding of the challenges faced by ML practitioners when developing ML models inside notebooks. The results indicate that ML practitioners work in a highly iterative fashion, often experimenting with multiple strategies to analyse the data or produce meaningful visualisations~\cite{kandel2012enterprise,kery2018story,liu2019understanding,chattopadhyay2020whats}. Studies have also been conducted to understand how practitioners generate, evaluate and manage alternative hypothesis, visual designs, methods, tools, algorithms and data sources to arrive at the final implementation~\cite{liu2019understanding,kandel2012enterprise}. The findings from these studies can be used as guidelines for improving existing notebook technologies or designing new ones.

To manage and prune multiple versions of code that accumulate over time when developing in notebooks, \emph{Head et al} propose a code gathering tool that allows practitioners to review and only keep the relevant version of code. Other tools such as \textit{WrangleDoc} and \textit{VizSmith} have also been proposed to aid ML practitioners when working within computational notebooks~\cite{yang2021subtle,bavishi2021vizsmith}.

@@

To test the correctness of ML models, several improvements over existing test adequacy metrics have been proposed. Tools such as \textit{DeepXplore} and \textit{DeepGauge} propose new test adequacy metrics such as neuron coverage adapted for ML enabled software systems~\cite{pei2017deepxplore,ma2018deepgauge,gerasimou2020importance-driven}. Formal verification methods have also been proposed that try to provide formal guarantee of robustness against adversarial examples. Such methods however are only feasible for statistical ML models and become computationally expensive for more complex models such as deep neural networks~\cite{zhu2021deepmemory,baluta2021scalable}. Several works have been conducted on generating and detecting adversarial inputs for ML models. Data augmentation techniques based on fuzzing, search based software testing and mutation testing have been proposed to generate adversarial examples that can be used during model training to improve its robustness~\cite{braiek2019deepevolution,gao2020fuzz,wang2021robot,zhang2020white-box}. It is however not possible to include all variations of adversarial examples into the training data. Thus, methods have been proposed to detect adversarial inputs during runtime~\cite{xiao2021self-checking,wang2020dissector,wang2019adversarial,berend2020cats}.

@@

\textit{ESCAPE}, \textit{GAM Coach}, \textit{Angler} and \textit{Drava} are a few other tools that have been proposed to handle specialised use-cases such as identifying systematic errors in ML models, generating counterfactual explanations for Generalised Additive Models, prioritising machine translation model improvements and relating human concepts with semantic dimensions extracted by ML models during disentangled representation learning\cite{ahn2023escape,wang2023gam,robertson2023angler,wang2023drava}.
